---
title: "soundDist_scratch"
output: html_document
date: "2024-01-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Write down the model for estimating source level from distance

# first try with lme4

```{r}
library(lme4)

install.packages("nimble")
library("nimble")
install.packages("baRulho")
library("baRulho")
```
# Model structure

## General


## Definitions

Indices:
* i: species
* j: event localized (a song that happened)
* k: clip (recording of song from known distance)

Constants applied to data before modeling
* d: distance of clip k's recorder from event j
* f: frequency of song recorded in event j


Sound parameters:
* `A_geo`: geometric spreading, 1/d**2
* `A_atm`: atmospheric loss, proportional to d and f**2
* `Ahab_ijk` = excess attenuation beyond spherical spreading and atmospheric attenuation, which we model as linear with habitat coefficients for open/closed (habitat_ijk) and landscape rugosity
* `L_i` = source volume for species i, dB SPL measurement of source volume, corrected by calibrating the ARU recording based on angle of arrival and by removing noise level

RSL of a clip is:

`RSL_ijk = L_i - A_geo - A_atm - Ahab_ijk`

We can calculate `A_geo` and `A_atm`, so subtract those from both sides and just model `Y`: under-estimated source volume based on physics:

`Y_ijk = RSL_ijk + A_geo(d) + A_atm(d) = L_i - Ahab_ijk`

## Model (Deterministic part only)

Underestimated source volume based on physics = Source volume - excess attenuation due to habitat:
* `Y_ijk = L_i - Ahab_ijk * f_i * d_ijk`

Divide everything by `(f_i * d_ijk)`:
* `Y_ijk/(f_i * d_ijk) = L_i/(f_i * d_ijk) - Ahab_ijk`

Redefine some things:
* `y_ijk = Y_ijk/(f_i * d_ijk)`
* `L*_i = L_i/(f_i * d_ijk)`

Add covariates for habitat attenuation:
* `Ahab_ijk = alpha_0 + alpha_1 * topography_ijk + alpha_2 * rugosity_ijk` 

Final model to put into LME:
* `y_ijk = (1|-1 + L*_i) - (alpha_0 + alpha_1 * topography_ijk + alpha_2 * rugosity_ijk)`



# 

```{markdown}
# Indices

# good explanation of nesting in lme4 
# https://www.muscardinus.be/statistics/nested.html#:~:text=Nested%20random%20effects%20assume%20that,relationship%20between%20the%20random%20effects.
# basically: if A is big grouping and B is sub-grouping of A, 
# random effects on A and B: (1|A) + (1|A:B) can be abbreviated into (1|A/B)
# and if B values are unique, rather than repeated, across A, 
# (1|A) + (1|B) is equivalent as well

# from the equation for observed sound level at distance d (Euclidean distance - 1m):
# L_obs = L_0 - A_geo - A_atm - A_hab
# A_hab = alpha_hab * freq * d
# A_geo = 20*log10(d)
# A_atm = ...
#
# where: 
# L_0: source level at 1 m, dB SPL
# d: distance to source - 1m
# A_geo: level reduction due to spherical spreading
# A_atm: loss due to atmospheric absorption
# A_hab: any remaining attenuation beyond A_geo and A_atm
# 
# isolate unkowns L_0 and A_hab=alpha_hab * f *d
# Y = L* + alpha_hab
# where Y is -1*(L_obs + A_geo + A_atm)/(freq*d)
# and L* is -1*L_0/(freq*d)
# 
# model right side as linear model, 
# with nested random effects on L* for species _and_ for event within species

clips = read.csv('../metadata/localized_clips_with_decibel.csv')

#calc Y from L_obs (corrected_dB_max), A_geo, A_atm
clips$freq <- (clips$low_f + clips$high_f)/2
r0 = 1 #1 meter reference point
clips$d = clips$distance_m - r0 #distance to "reference" d0=1m
ref_pressure = 20e-6 # 20 micro Pascals, reference of dB SPL in air
# level_pascals = 10^(clips$corrected_dB_max/20) * (ref_pressure)^2 
atm_atten = atmospheric_attenuation(
  clips$freq,clips$temp, RH=0.7, p=101325,
  formula = 1)
clips$A_atm <- atm_atten$atmospheric.attenuation * clips$d

# geometric spreading
clips$A_geo <- 20 * log10(clips$distance_m/r0)

clips$Y <- -1*(clips$corrected_dB_max + clips$A_geo + clips$A_atm)/(clips$d*clips$freq)
# note that the random effect on event id is on L* = -1*L_0/(d*freq)

# remaining q: is L* eestimated here, or do we need to include it somewhere?
# I think that the random effect estimates N(mu,sigma) not just N(0,sigma) so its included in the random effect
# but check w/Justin/Tessa
# also, we probably want a _fixed_ effect for species? otherwise we don't estimate a per-species L0
# m1 <- lmer(Y ~ 1 + (1 | class_name) + (1 | class_name:event_id), data = clips)  # (1| class_name) + (1 | event_id)

# so perhaps: fixed effect for class name, random effect for event_id 
# we are not fitting a sigma for species volume; 
m0 <- lm(Y ~ 1 + class_name , data = clips) 

L0 = -fixedeffect * f * d
# alpha0 is the intercept

# back-calculate L0_i per species and 
m0 <- lmer(Y ~ 1  + class_name + (1 | event_id), data = clips) 
m1 <- lmer(Y ~ 1  + habitat_closed + class_name + (1 | event_id), data = clips) 
m2 <- lmer(Y ~ 1  + habitat_rugosity + habitat_closed + class_name + (1 | event_id), data = clips) 

# add covariates:
#+ a_rugosity*habitat_rugosity + a_rugosity*habitat_closed
m1
```



```{r}

# need to make nested lists of knowns: Y[i,j,k], f[i], d[i,j,k], habitat[i,j,k], rugosity[i,j,k]


modelCode <- nimbleCode({
  #model
  for (i in 1:N_SPECIES){
    for (j in 1:N_EVENTS[i]){
      for (k in 1:N_CLIPS[i,j]){
        Y[i,j,k] = L[i] - Ahab[i,j,k] * f[i] * d[i,j,k]
        Ahab_ijk ~ dnorm(muH[i,j,k],sigmaH)
        muH[i,j,k] = alpha_0 + alpha_1 * habitat[i,j,k] + alpha_2 * rugosity[i,j,k]
        L[i] ~ dnorm(muL,sigmaL)
      }
    }
  }
  
  # priors
  # L[i] sigmaH, alpha_0, alpha_1, alpha_2
  
  
})
```


```{r}
```

