---
title: "soundDist_scratch"
output: html_document
date: "2024-01-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Model structure

### General

# from the equation for observed sound level at distance d (Euclidean distance - 1m):
# L_obs = L_0 - A_geo - A_atm - A_hab
# A_hab = alpha_hab * freq * d
# A_geo = 20*log10(d)
# A_atm = ...
#
# where: 
# L_0: source level at 1 m, dB SPL
# d: distance to source - 1m
# A_geo: level reduction due to spherical spreading
# A_atm: loss due to atmospheric absorption
# A_hab: any remaining attenuation beyond A_geo and A_atm
# 

### Definitions

Indices:
* i: species
* j: event localized (a song that happened)
* k: clip (recording of song from known distance)

Constants applied to data before modeling
* d: distance of clip k's recorder from event j
* f: frequency of song recorded in event j

Sound parameters:
* `A_geo_ijk`: geometric spreading, 1/d_ijk**2
* `A_atm_ijk`: atmospheric loss, proportional to d_ijk and f_i**2
* `Ahab_ijk` = excess attenuation beyond spherical spreading and atmospheric attenuation, which we model as linear with habitat coefficients for open/closed (habitat_ijk) and landscape rugosity
* `L_i` = true sound level for species i, dB SPL measurement of source volume, corrected by calibrating the ARU recording based on angle of arrival and by removing noise level

RSL of a clip is:

`RSL_ijk := L_i - A_geo - A_atm - Ahab_ijk`

We can calculate `A_geo` and `A_atm`, so add those to both sides and just model `Y`: under-estimated source volume based on physics:

`RSL_ijk + A_geo(d_ijk) + A_atm(d_ijk, f_i) = L_i - Ahab(d_ijk, f_i)`

Define the quantity on the left side as `Y_ijk`, the underestimated source volume based on physics = Source volume - excess attenuation due to habitat: 

* `Y_ijk := RSL_ijk + A_geo(d_ijk) + A_atm(d_ijk, f_i)`

Also we define `Ahab` as:
* `Ahab(d_ijk, f_i) := alphahab_ijk * f_i * d_ijk`

So we have finally:
* `Y_ijk = L_i - alphahab_ijk * f_i * d_ijk`

### Model (Deterministic part only)

Divide everything by `(-f_i * d_ijk)`:
* `Y_ijk/(-f_i * d_ijk) = L_i/(-f_i * d_ijk) + alphahab_ijk`

Redefine some things:
* `Y*_ijk := Y_ijk/(-f_i * d_ijk)`
* `L*_i := L_i/(-f_i * d_ijk)`

So then we have:
* `Y*_ijk = L*_i + alphahab_ijk`

Add covariates for habitat attenuation:
* `alphahab_ijk := alpha_0 + alpha_1 * habitat_tree_cover` 

Example of a final model to put into LME:
* `Y*_ijk = L*_i + alpha_0 + alpha_1 * habitat_tree_cover`

## Data transformations

### To put into model

We have the calibrated RSL (noise removed, recorder calibration accounted for)

Turn it into the modeled quantity `y_ijk`

```
Y*_ijk = Y_ijk / (-f_i * d_ijk)
      = (RSL_ijk + A_geo(d) + A_atm(d)) / (-f_i * d_ijk)
```

### To get out of model

What is the true `L_i` for each species at distance = 1?

In our case we're claiming that you have to add alpha_0 to the species covariate to get the true L_i?

SO the covariate gotten from the model for the species, plus the offset for the intercept (which is alpha_0) is:

```
Y*_ijk = L*_ijk + alpha_0 + alpha_1 * habitat_tree_cover_ijk

# Replace the stars
Y_ijk / (-f_i * d_ijk) = L_ijk/(-f_i * d_ijk) + alpha_0 + alpha_1 * habitat_tree_cover_ijk

# Replace Y_ijk
(RSL_ijk + A_geo(d_ijk) + A_atm(d_ijk, f_i)) / (-f_i * d_ijk) = L_ijk/(-f_i * d_ijk) + alpha_0 + alpha_1 * habitat_tree_cover_ijk

#Switch the L_ijk term and the RSL term sides
L_ijk/(f_i * d_ijk) = (RSL_ijk + A_geo(d_ijk) + A_atm(d_ijk, f_i)) / (f_i * d_ijk) + alpha_0 + alpha_1 * habitat_tree_cover_ijk

# Now we get
L_i = RSL_ijk + A_geo(d_ijk) + A_atm(d_ijk, f_i) + (alpha_0 + alpha_1 * habitat_tree_cover_ijk)*(f_i * d_ijk)

```


```{r}
intercept <- -0.0002688748
habitat_tree_cover_covariate <- -0.0012270960
habitat_tree_cover_mean <- 0.722
class_covariate <- 0.0007018519 #hermit thrush
f_i <- 3500#of hermit thrush
d_ijk <- 1 #(get 1m)

L_i = -(class_covariate + intercept)*(-f_i * d_ijk)
```


# isolate unkowns L_0 and A_hab=alpha_hab * f *d
# Y = L* + alpha_hab
# where Y is -1*(L_obs + A_geo + A_atm)/(freq*d)
# and L* is -1*L_0/(freq*d)
# 
# model right side as linear model, 
# with nested random effects on L* for species _and_ for event within species


## Write down the model for estimating source level from distance

# first try with lme4

```{r}
library(lme4)

install.packages("nimble")
library("nimble")
install.packages("baRulho")
library("baRulho")
```


```{markdown}
# Indices

# good explanation of nesting in lme4 
# https://www.muscardinus.be/statistics/nested.html#:~:text=Nested%20random%20effects%20assume%20that,relationship%20between%20the%20random%20effects.
# basically: if A is big grouping and B is sub-grouping of A, 
# random effects on A and B: (1|A) + (1|A:B) can be abbreviated into (1|A/B)
# and if B values are unique, rather than repeated, across A, 
# (1|A) + (1|B) is equivalent as well

# from the equation for observed sound level at distance d (Euclidean distance - 1m):
# L_obs = L_0 - A_geo - A_atm - A_hab
# A_hab = alpha_hab * freq * d
# A_geo = 20*log10(d)
# A_atm = ...
#
# where: 
# L_0: source level at 1 m, dB SPL
# d: distance to source - 1m
# A_geo: level reduction due to spherical spreading
# A_atm: loss due to atmospheric absorption
# A_hab: any remaining attenuation beyond A_geo and A_atm
# 
# isolate unkowns L_0 and A_hab=alpha_hab * f *d
# Y = L* + alpha_hab
# where Y is -1*(L_obs + A_geo + A_atm)/(freq*d)
# and L* is -1*L_0/(freq*d)
# 
# model right side as linear model, 
# with nested random effects on L* for species _and_ for event within species
```

```{r}
clips = read.csv('../metadata/localized_clips_with_decibel.csv')
clips$class_name <- factor(clips$class_name)
clips$rel_hum <- clips$rel_hum / 100
#calc Y from L_obs (corrected_dB_max), A_geo, A_atm
clips$freq <- (clips$low_f + clips$high_f)/2
r0 = 1 #1 meter reference point
clips$d = clips$distance_m - r0 #distance to "reference" d0=1m

# PA data has temp from audiomoth
clips$temp_C <- ifelse(!is.na(clips$temp), clips$temp, clips$temp_C)

# ref_pressure = 20e-6 # 20 micro Pascals, reference of dB SPL in air
# level_pascals = 10^(clips$corrected_dB_max/20) * (ref_pressure)^2 
atm_atten = atmospheric_attenuation(
  clips$freq,
  clips$temp_C, 
  RH=clips$rel_hum, 
  p=101325,
  formula = 2
) # formula 2: 2007 algorithm; units are dB/meter
# atmospheric absoprtion is atm_atten coef * distance 
clips$A_atm <- atm_atten$atmospheric.attenuation * clips$d

# geometric spreading
clips$A_geo <- 20 * log10(clips$distance_m/r0)

clips$Y <- -1*(clips$corrected_dB_max + clips$A_geo + clips$A_atm)/(clips$d*clips$freq)
# note that the random effect on event id is on L* = -1*L_0/(d*freq)

# remaining q: is L* eestimated here, or do we need to include it somewhere?
# I think that the random effect estimates N(mu,sigma) not just N(0,sigma) so its included in the random effect
# but check w/Justin/Tessa
# also, we probably want a _fixed_ effect for species? otherwise we don't estimate a per-species L0
# m1 <- lmer(Y ~ 1 + (1 | class_name) + (1 | class_name:event_id), data = clips)  # (1| class_name) + (1 | event_id)

# filter out rows without Y values

# so perhaps: fixed effect for class name, random effect for event_id 
# we are not fitting a sigma for species volume; 
m0 <- lm(Y ~ 1 + class_name , data = clips) 

# L0 = -fixedeffect * f * d
# alpha0 is the intercept

# remove rows if 
# clips <- clips %>%
#   na.omit(corrected_dB_max)

# Save these calculated values to a new CSV
write.csv(clips, "../metadata/prediction_df.csv", row.names=FALSE)

# back-calculate L0_i per species and A_hab(f,covars)
m0 <- lmer(Y ~ 1  + class_name + (1 | event_id), data = clips) 
m1 <- lmer(Y ~ 1  + habitat_tree_cover + class_name + (1 | event_id), data = clips) 
m2 <- lmer(Y ~ 1 + habitat_tree_cover + class_name + (1 | event_id), data = clips) 
m3 <- lm(Y ~ 1 + habitat_tree_cover + class_name, data = clips) 

clips$preds_m3 <- predict(m3)
# levels(preds$class_name)
# preds$m3_pred_Y <- predict(m3,newdata=preds)

# Assuming your dataframe is named 'your_dataframe'
library(dplyr)


clips_no_rare_classes <- clips %>%
  group_by(class_name) %>%
  filter(n() >= 50) %>%
  ungroup()

# clips_no_sparse_events <- clips_no_rare_classes %>%
#   group_by(event_id) %>%
#   filter(n() >= 3) %>%
#   ungroup()

#all of the models with event_id random effect are showing 
#"boundary (singular) fit: see help('isSingular')"
m4 <- lmer(Y ~ 1 + habitat_tree_cover + class_name + (1 | event_id), data = clips_no_rare_classes) 

# add covariates:
#+ a_rugosity*habitat_rugosity + a_rugosity*habitat_closed
summary(m3)
```

```{r}
predict(m3)
```




```{r}

# need to make nested lists of knowns: Y[i,j,k], f[i], d[i,j,k], habitat[i,j,k], rugosity[i,j,k]


modelCode <- nimbleCode({
  #model
  for (i in 1:N_SPECIES){
    for (j in 1:N_EVENTS[i]){
      for (k in 1:N_CLIPS[i,j]){
        Y[i,j,k] = L[i] - Ahab[i,j,k] * f[i] * d[i,j,k]
        Ahab_ijk ~ dnorm(muH[i,j,k],sigmaH)
        muH[i,j,k] = alpha_0 + alpha_1 * habitat[i,j,k] + alpha_2 * rugosity[i,j,k]
        L[i] ~ dnorm(muL,sigmaL)
      }
    }
  }
  
  # priors
  # L[i] sigmaH, alpha_0, alpha_1, alpha_2
  
  
})
```


```{r}
```

